Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-07 paper.tex
more wake/sleep

Inference in the generative model in Figure~\ref{graphicalModel}(a) is
complicated by the fact that the programs are unobserved,
and so we must solve a difficult search problem to find any programs at all. To make
search tractable we learn a bottom-up \emph{recognition
  model}, written $q(\cdot )$
(Figure~\ref{graphicalModel}b).  The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference,ritchie2016deep} for the generative
model.

@@

Our goal  is to induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) a \emph{wake} phase where we search for programs, and then a pair of \emph{sleep} phases: (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that these three steps bootstrap off each other:

@@

%% The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
%% learning a generative model and training a recognition model on samples from the generative model.
%% We apply this idea to program induction, motivating why we call our algorithm the \system.
Figure~\ref{feeding} diagrams how these three steps feed off of each other.
\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}
Mechanistically, \system does approximate MAP inference in the generative model of Figure~\ref{graphicalModel}(a) by maximizing this lower bound on the joint probability.
We maximize $\lowerBound$ by alternatingly maximizing w.r.t. the frontiers and the DSL:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $\lowerBound$ w.r.t. the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $\int \lowerBound\;\mathrm{d}\theta$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound\;\mathrm{d}\theta$.
Once we have a DSL $\mathcal{D}$ we can update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by learning a neural recognition model:

@@

\subsection{Searching for Programs}\label{explorationSection}

@@

\noindent\textbf{Definition.}
A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
\\\noindent\textbf{Definition.}
A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
one number for each DSL primitive $e:\tau\in \mathcal{D}$, written $\theta_e$,
and a weight controlling the probability of a variable occurring in a program, written $\theta_{\text{var}}$.

Algorithm~\ref{programGenerativeModel} is a procedure for drawing
samples from $\probability[p|\mathcal{D},\theta]$.  In practice, we
enumerate programs rather than sampling them.  Enumeration proceeds by
a depth-first search over the random choices made by
Algorithm~\ref{programGenerativeModel}; we wrap the depth-first search
in iterative deepening to (approximately) build $\lambda$-calculus
expressions in order of their probability.

@@

\subsection{Learning a Neural Recognition Model}\label{recognitionSection}

@@

\subsection{Inducing the DSL from the Frontiers}\label{grammarInductionSection}

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at the programs in the frontiers and
generalize beyond them; 
not only so the DSL can better express the current solutions,
but also so that the DSL might expose new abstractions
which will later be used to
discover even better programs.

>>>>

Inference in the generative model in Figure~\ref{graphicalModel}(a) is
complicated by the fact that the programs are unobserved,
and so we must solve a difficult search problem to find any programs at all. To make
search tractable we learn a bottom-up \emph{recognition
  model}, written $q(\cdot )$
(Figure~\ref{graphicalModel}b).  The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. $q(\cdot )$ implements an amortized
inference
scheme~\cite{le2016inference} for the generative %ritchie2016deep
model.

@@

Our goal  is to induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) a \emph{wake} phase where we search for programs, and then a pair of \emph{sleep} phases: (2) learning a neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that these three steps bootstrap off each other (Figure~\ref{feeding}):

@@

%% The Helmholtz machine~\cite{dayan1995helmholtz} introduced the idea of alternating between
%% learning a generative model and training a recognition model on samples from the generative model.
%% We apply this idea to program induction, motivating why we call our algorithm the \system.

\begin{figure}\centering
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs: $p$};
    \node[align = center](DSL) at (3,1) {DSL: $\mathcal{D}$};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model: $q$};

@@

%% In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
%% Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
%% \begin{align}
%%   A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
%% &+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
%%   \end{align}
\system does approximate MAP inference in the generative model of Figure~\ref{graphicalModel}(a) by maximizing this lower bound on the joint probability,
alternatingly maximizing w.r.t. the frontiers (``wake'') and the DSL (``sleep''):
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $\lowerBound$ w.r.t. the frontiers.} Here $(\mathcal{D},\theta)$ is fixed and we
want to find new programs to add to  the frontiers so that $\lowerBound$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $\int \lowerBound\;\mathrm{d}\theta$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed so we can evaluate $\lowerBound$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $\int \lowerBound\;\mathrm{d}\theta$.
Once we have a DSL $\mathcal{D}$ we can update $\theta$ to $\argmax_\theta \lowerBound(\mathcal{D},\theta,\left\{\mathcal{F}_x \right\})$.

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by learning a neural recognition model
during a second ``sleep'' phase:

@@

\subsection{Wake Phase: Searching for Programs}\label{explorationSection}

@@

\noindent\textbf{Definition: $(\mathcal{D},\theta)$.}
A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
%\\\noindent\textbf{Definition.}
A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
one number for each DSL primitive $e:\tau\in \mathcal{D}$, written $\theta_e$,
and a weight controlling the probability of a variable occurring in a program, $\theta_{\text{var}}$.

Algorithm~\ref{programGenerativeModel} is a procedure for drawing
samples from $\probability[p|\mathcal{D},\theta]$.  In practice, we
enumerate programs rather than sampling them.  Enumeration proceeds by
a depth-first search over the random choices made by
Algorithm~\ref{programGenerativeModel}; we wrap the depth-first search
in iterative deepening to build $\lambda$-calculus
expressions in order of their probability.

@@

\subsection{Sleep Phase I: Learning a Recognition Model}\label{recognitionSection}

@@

\subsection{Sleep Phase II: Inducing the DSL}\label{grammarInductionSection}

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at  the frontiers and
generalize beyond them, 
both so the DSL can better express the current solutions,
and  also so that the DSL might expose new abstractions
which will later be used to
discover more programs.
